---
title: "Presentation ready plots"
author: "MACS 40700 <br /> University of Chicago"
output: rcfss::xaringan_wide
---

```{r setup, include = FALSE, cache = FALSE}
# generate CSS file
library(xaringanthemer)
rcfss::xaringan_theme()

# source in the default knitr options
source(here::here("R", "slide-opts.R"))

# enable panelsets and default theme
xaringanExtra::use_panelset()
# ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# load basic packages
library(knitr)
library(here)
library(countdown)
library(patchwork)
```

class: middle, inverse

# Welcome

---

## Announcements I

Project presentations on Thursday

- Bring a laptop to fill out feedback forms
- Presentation order, with latest start-by times:
  - 11:03 - Calm Comet
  - 11:11 - Chalky Piglet
  - 11:19 - Elated Anura
  - 11:27 - Female Python
  - 11:35 - Froggy Panda
  - 11:43 - Outer Baboon
  - 12:51 - Sleepy Spider
  - 11:59 - Swampy Pygmy
  - 12:07 - Tan Egg
  - 12:15 - Undead Hen

---

## Setup

```{r message = FALSE, warning = FALSE}
# load packages
library(tidyverse)
library(colorspace)
library(ggtext)
library(scales)
library(ggrepel)

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, fig.retina = 3,
  dpi = 300, out.width = "60%"
)

# dplyr print min and max
options(dplyr.print_max = 6, dplyr.print_min = 6)
```

---

## Agenda for today

- Wrap up annotations

- Telling a story

- `xaringan`

- Plots on slides

---

class: inverse, middle

# Telling a story

---

## Multiple ways of telling a story

- Sequential plots: Motivation, then resolution

- A single plot: Resolution, and hidden in it motivation

<br>

.note[
**Project note:** you're asked to create two plots per question. One possible approach: Start with a plot showing the raw data, and show derived quantities (e.g. percent increases, averages, coefficients of fitted models) in the subsequent plot.
]

---

## Simplicity vs. complexity

> When you’re trying to show too much data at once you may end up not showing anything.

- Never assume your audience can rapidly process complex visual displays

- Don't add variables to your plot that are tangential to your story

- Don't jump straight to a highly complex figure; first show an easily digestible subset (e.g., show one facet first)

- Aim for memorable, but clear

<br>

.note[
**Project note:** Make sure to leave time to iterate on your plots after you practice your presentation. If certain plots are getting too wordy to explain, take time to simplify them!
]

---

## Consistency vs. repetitiveness

> Be consistent but don’t be repetitive.

- Use consistent features throughout plots (e.g., same color represents same level on all plots)

- Aim to use a different type of visualization for each distinct analysis

<br>

.note[
**Project note:** If possible, ask a friend who is not in the class to listen to your presentation and then ask them what they remember. Then, ask yourself: is that what you wanted them to remember?
]

---

class: inverse, middle

# `xaringan`

---

## `xaringan`

- R Markdown extension to HTML5 presentations

- [Detailed guide to `xaringan`](https://arm.rbind.io/slides/xaringan.html)

- [`xaringan` in R Markdown reference book](https://bookdown.org/yihui/rmarkdown/xaringan.html)

- [Example slides](https://dataviz-sp22.github.io/proj1/presentation/presentation.html#1)

---

class: inverse, middle

# Plots on slides

---

## Fitting your figures on a slide

- Use a setup chunk for consistent `dpi`, `fig.asp`, `out.width` etc.

- But, remember that you might want different `fig.asp` and `out.width` for faceted plots

- For a "zoomed out" look, increase `fig.width`

.pull-left[
```{r out.width = "100%", echo = FALSE}
mtcars %>%
  rownames_to_column() %>%
  ggplot(aes(x = disp, y = mpg)) +
  geom_text_repel(aes(label = rowname)) +
  coord_cartesian(clip = "off") +
  labs(title = "fig.width = 8")
```
]
.pull-right[
```{r out.width = "100%", fig.width = 10, echo = FALSE}
mtcars %>%
  rownames_to_column() %>%
  ggplot(aes(x = disp, y = mpg)) +
  geom_text_repel(aes(label = rowname)) +
  coord_cartesian(clip = "off") +
  labs(title = "fig.width = 10")
```
]

---

## Multiple figures on a slide

- Use `.pull-left` and `.pull-right` environments (example in slide template)

- Use `out.width= "50%"` in a chunk that creates two plots to have each span half the slide and be placed next to each other (or `out.width = "33%"` for three plots, etc.)

- Use the [**patchwork**](https://patchwork.data-imaginist.com/) package (details on this later in the semester, but feel free to look into it yourself!)

- Possibly, use pivoting to reshape your data and then use facets

---

## Like something you saw in my slides?

<br>

Look into the source code at https://github.com/uc-dataviz/slides.
