---
title: "Deep dive: geoms"
author: "MACS 40700 <br /> University of Chicago"
output: rcfss::xaringan_wide
---

```{r setup, include = FALSE, cache = FALSE}
# generate CSS file
library(xaringanthemer)
rcfss::xaringan_theme()

# source in the default knitr options
source(here::here("R", "slide-opts.R"))

# enable panelsets and default theme
xaringanExtra::use_panelset()
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# make out width 60% to accommodate panelsets
knitr::opts_chunk$set(
  out.width = "60%"
)

# load basic packages
library(knitr)
library(here)
library(countdown)
```

class: middle, inverse

# Announcements

---

## Announcements

- HW 1 is posted:
  - Go to the course GitHub org and look for a repo named `homework-01-YOUR_GITHUB_NAME`
  - Instructions are in the README
  - All work must be completed in `homework-01.Rmd` document, knit, and submitted along with the output `homework-01.md`
  - Due Tuesday, Apr 12, by 11am
  - At the deadline you'll lose push access to your repo, if you want to submit late, email me to enable late submission
  - Post questions on the discussion forum

- Check email before class on Tuesday for your team assignments, sit with your teams in class on Tuesday

---

class: middle, inverse

# Setup

---

## Packages

```{r}
# load packages
library(tidyverse)
# remotes::install_github("uc-cfss/rcfss)
library(rcfss)
```

---

## ggplot2 theme

```{r}
# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

---

## Figure sizing

For more on including figures in R Markdown documents with the right size, resolution, etc. the following resources are great:

- [R for Data Science - Graphics for communication](https://r4ds.had.co.nz/graphics-for-communication.html)
- [Tips and tricks for working with images and figures in R Markdown documents](https://www.zevross.com/blog/2017/06/19/tips-and-tricks-for-working-with-images-and-figures-in-r-markdown-documents/)

```{r eval = FALSE}
# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8, # 8"
  fig.asp = 0.618, # the golden ratio
  fig.retina = 3, # dpi multiplier for displaying HTML output on retina
  dpi = 300, # higher dpi, sharper image
  out.width = "70%"
)
```

---

## Data prep: new variables

From last class...

```{r}
scorecard <- scorecard %>%
  mutate(pctpell_cat = cut_interval(x = pctpell, n = 6)) %>%
  drop_na(pctpell_cat)

scorecard %>%
  count(pctpell_cat)
```

---

## Data prep: summary table

From last class...

```{r}
mean_netcost_pctpell <- scorecard %>%
  group_by(pctpell_cat) %>%
  summarise(mean_netcost = mean(netcost, na.rm = TRUE))

mean_netcost_pctpell
```

---

class: middle, inverse

# Aesthetic mappings in ggplot2

---

## Global vs. layer-specific aesthetics

- Aesthetic mappings can be supplied in the initial `ggplot()` call, in individual layers, or in some combination of both.

- Within each layer, you can add, override, or remove mappings.

- If you only have one layer in the plot, the way you specify aesthetics doesn’t make any difference. However, the distinction is important when you start adding additional layers.

---

## Activity: Spot the difference I

.task[
Do you expect the following plots to be the same or different? If different, how? Discuss in a pair (or group) without running the code and sketch the resulting plots based on what you think the code will produce.
]

```{r fig.show = "hide"}
# Plot A
ggplot(scorecard, aes(x = satavg, y = netcost, color = pctpell_cat)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5)

# Plot B
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point(aes(color = pctpell_cat), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5)
```

```{r echo = FALSE, cache = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## Activity: Spot the difference II

.task[
Do you expect the following plots to be the same or different? If different, how? Discuss in a pair (or group) without running the code and sketch the resulting plots based on what you think the code will produce.
]

```{r fig.show = "hide"}
# Plot A
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point(aes(color = pctpell_cat))

# Plot B
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point(color = "blue")

# Plot C
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point(color = "#a493ba")
```

```{r echo = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

class: middle, inverse

# Geoms

---

## Geoms

- Geometric objects, or geoms for short, perform the actual rendering of the layer, controlling the type of plot that you create

- You can think of them as "the geometric shape used to represent the data"

---

## One variable

- Discrete:
  - `geom_bar()`: display distribution of discrete variable.

- Continuous
  - `geom_histogram()`: bin and count continuous variable, display with bars
  - `geom_density()`: smoothed density estimate
  - `geom_dotplot()`: stack individual points into a dot plot
  - `geom_freqpoly()`: bin and count continuous variable, display with lines

---

## .hand[aside...]

Always use "typewriter text" (monospace font) when writing function names, and follow with `()`, e.g., 

- `geom_freqpoly()`
- `mean()`
- `lm()`

---

## `geom_bar()`

```{r}
ggplot(scorecard, aes(x = pctpell_cat)) +
  geom_bar()
```

---

## `geom_bar()`

```{r}
ggplot(scorecard, aes(y = pctpell_cat)) +
  geom_bar()
```

---

## `geom_histogram()`

```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_histogram()
```

---

## `geom_histogram()` and `binwidth`

.panelset[
.panel[.panel-name[2K]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_histogram(binwidth = 2000)
```
]
.panel[.panel-name[5K]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_histogram(binwidth = 5000)
```
]
.panel[.panel-name[10K]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_histogram(binwidth = 10000)
```
]
.panel[.panel-name[20K]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_histogram(binwidth = 20000)
```
]
]

---

## `geom_density()`

```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density()
```

---

## `geom_density()` and bandwidth (`bw`)

.panelset[
.panel[.panel-name[1]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(bw = 1)
```
]
.panel[.panel-name[100]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(bw = 100)
```
]
.panel[.panel-name[1000]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(bw = 1000)
```
]
.panel[.panel-name[10000]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(bw = 10000)
```
]
]

---

## `geom_density()` outlines

.panelset[
.panel[.panel-name[full]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(outline.type = "full")
```
]
.panel[.panel-name[both]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(outline.type = "both")
```
]
.panel[.panel-name[upper]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(outline.type = "upper")
```
]
.panel[.panel-name[lower]
```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_density(outline.type = "lower")
```
]
]

---

## `geom_dotplot()`

.task[
What does each point represent? How are their locations determined? What do the x and y axes represent?
]

```{r fig.asp = 0.4, fig.width = 12, out.width = "80%"}
ggplot(scorecard, aes(x = netcost)) +
  geom_dotplot(binwidth = 418)
```

```{r echo = FALSE, cache = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## `geom_freqpoly()`

```{r}
ggplot(scorecard, aes(x = netcost)) +
  geom_freqpoly(binwidth = 1000)
```

---

## `geom_freqpoly()` for comparisons

.panelset[
.panel[.panel-name[Histogram]
```{r}
ggplot(scorecard, aes(x = netcost, fill = pctpell_cat)) +
  geom_histogram(binwidth = 5000)
```
]
.panel[.panel-name[Frequency polygon]
```{r}
ggplot(scorecard, aes(x = netcost, color = pctpell_cat)) +
  geom_freqpoly(binwidth = 5000, size = 1)
```
]
]

---

## Two variables - both continuous

- `geom_point()`: scatterplot
- `geom_quantile()`: smoothed quantile regression
- `geom_rug()`: marginal rug plots
- `geom_smooth()`: smoothed line of best fit
- `geom_text()`: text labels

---

## `geom_rug()`

```{r}
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point() +
  geom_rug()
```

---

## `geom_rug()` on the outside

```{r}
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point() +
  geom_rug(outside = TRUE) +
  coord_cartesian(clip = "off")
```

---

## `geom_rug()` on the outside, but better

```{r}
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point() +
  geom_rug(outside = TRUE, sides = "tr") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(1, 1, 1, 1, "cm"))
```

---

## `geom_text()`

```{r}
sample_n(scorecard, size = 50) %>%
  ggplot(aes(x = satavg, y = netcost)) +
  geom_text(aes(label = type))
```


---

## `geom_text()` and more

```{r}
sample_n(scorecard, size = 50) %>%
  ggplot(aes(x = satavg, y = netcost)) +
  geom_text(aes(label = type, color = type))
```

---

## `geom_text()` and even more

```{r}
sample_n(scorecard, size = 50) %>%
  ggplot(aes(x = satavg, y = netcost)) +
  geom_text(aes(label = type, color = type),
    show.legend = FALSE
  )
```

---

## Two variables - show distribution

- `geom_bin2d()`: bin into rectangles and count
- `geom_density2d()`: smoothed 2d density estimate
- `geom_hex()`: bin into hexagons and count

---

## `geom_hex()`

```{r}
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_hex()
```

---

## `geom_hex()` and warnings

- Requires installing the [**hexbin**](https://cran.r-project.org/web/packages/hexbin/index.html) package separately!

```{r eval = FALSE}
install.packages("hexbin")
```

- Otherwise you might see

```
Warning: Computation failed in `stat_binhex()`
```

---

## Two variables

- At least one discrete
  - `geom_count()`: count number of point at distinct locations
  - `geom_jitter()`: randomly jitter overlapping points

- One continuous, one discrete
  - `geom_col()`: a bar chart of pre-computed summaries
  - `geom_boxplot()`: boxplots
  - `geom_violin()`: show density of values in each group

---

## `geom_jitter()`

.task[
How are the following three plots different?
]

.panelset[
.panel[.panel-name[Plot A]
```{r fig.asp = 0.4}
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_point()
```
]
.panel[.panel-name[Plot B]
```{r fig.asp = 0.4}
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_jitter()
```
]
.panel[.panel-name[Plot C]
```{r fig.asp = 0.4}
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_jitter()
```
]
]

```{r echo = FALSE, cache = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## `geom_jitter()` and `set.seed()`

.panelset[
.panel[.panel-name[Plot A]
```{r fig.asp = 0.5}
set.seed(123)
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_jitter()
```
]
.panel[.panel-name[Plot B]
```{r fig.asp = 0.5}
set.seed(123)
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_jitter()
```
]
.panel[.panel-name[Plot C]
```{r fig.asp = 0.5}
set.seed(234)
ggplot(scorecard, aes(x = type, y = netcost)) +
  geom_jitter()
```
]
]

---

## Two variables

- One time, one continuous
  - `geom_area()`: area plot
  - `geom_line()`: line plot
  - `geom_step()`: step plot

- Display uncertainty:
  - `geom_crossbar()`: vertical bar with center
  - `geom_errorbar()`: error bars
  - `geom_linerange()`: vertical line
  - `geom_pointrange()`: vertical line with center

- Spatial
  - `geom_sf()`: for map data (more on this later...)

---

## Average net cost per Pell grant recipient

```{r}
mean_netcost_pctpell <- scorecard %>%
  mutate(pctpell = round(pctpell, digits = 2)) %>%
  group_by(pctpell) %>%
  summarise(
    n = n(),
    mean_netcost = mean(netcost),
    sd_netcost = sd(netcost)
  ) %>%
  drop_na(mean_netcost)

mean_netcost_pctpell
```

---

## `geom_line()`

```{r}
ggplot(mean_netcost_pctpell, aes(x = pctpell, y = mean_netcost)) +
  geom_line()
```

---

## `geom_area()`

```{r}
ggplot(mean_netcost_pctpell, aes(x = pctpell, y = mean_netcost)) +
  geom_area()
```

---

## `geom_step()`

```{r}
ggplot(mean_netcost_pctpell, aes(x = pctpell, y = mean_netcost)) +
  geom_step()
```

---

## `geom_errorbar()`

.task[
Describe how this plot is constructed and what the points and the lines (error bars) correspond to.
]

.panelset[
.panel[.panel-name[Code]
```{r errorbar, fig.show = "hide"}
ggplot(mean_netcost_pctpell, aes(x = pctpell, y = mean_netcost)) +
  geom_point() +
  geom_errorbar(aes(
    ymin = mean_netcost - sd_netcost,
    ymax = mean_netcost + sd_netcost
  ))
```
]
.panel[.panel-name[Plot]
```{r ref.label = "errorbar", echo = FALSE}

```
]
]

```{r echo = FALSE, cache = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## Let's clean things up a bit!

Meet your new best friend, the [**scales**](https://scales.r-lib.org/) package!

```{r}
library(scales)
```

---

## Let's clean things up a bit!

.panelset[
```{r panelset = c(output = "Plot", source = "Code"), fig.show = "hold"}
ggplot(scorecard, aes(x = satavg, y = netcost)) +
  geom_point(alpha = 0.4, size = 2, color = "#012169") +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    x = "Average SAT score",
    y = "Net cost of attendance (USD)",
    title = "Test scores and net cost of attendance in US universities",
    subtitle = "2018-19",
    caption = "Source: College Scorecard"
  )
```
]

---

class: middle, inverse

# geoms

---

## Three variables

- `geom_tile()`: tile the plane with rectangles
- `geom_raster()`: fast version of `geom_tile()` for equal sized tiles

---

## `geom_tile()`

```{r}
ggplot(scorecard, aes(x = type, y = locale)) +
 geom_tile(aes(fill = netcost))
```

---

## Another look at smooth-ish surface

.panelset[
.panel[.panel-name[Summarize]
.small[
```{r}
mean_netcost_type_locale <- scorecard %>%
  group_by(type, locale) %>%
  summarize(mean_netcost = mean(netcost, na.rm = TRUE), .groups = "drop")

mean_netcost_type_locale
```
]
]
.panel[.panel-name[Plot]
```{r}
ggplot(mean_netcost_type_locale, aes(x = type, y = locale)) +
  geom_point(aes(color = mean_netcost), size = 40, pch = "square")
```
]
]

---

## Activity: Pick a geom

.task[
For each of the following problems, suggest a useful geom:

1. Display how the value of variable has changed over time
1. Show the detailed distribution of a single continuous variable
1. Focus attention on the overall relationship between two variables in a large dataset
1. Label outlying points in a single variable
]

```{r echo = FALSE, cache = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

class: inverse, middle

# Homework 1

---

## Homework 1

.task[
- Locate your Homework 1 repo, clone it in RStudio
- Put your name in the YAML of `homework-01.Rmd`, knit, commit (with a meaningful message), push
- Skim the README to get a sense of the instructions and use the remaining time to ask any clarifying questions and to get started on the first question
- Make sure you know how to ask a question on the course discussion forum, and set up notifications
]

---

## Homework 1 tips

- You can go in order or jump around between exercises. If you get stuck
  on a specific exercise, feel free to switch to a different one.

- But, of course, do what feels right for you in terms of pace, ordering, etc.!

- And ask questions on the course discussion forum
