---
title: "Interactive visualization"
author: "MACS 40700 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE, cache = FALSE}
# generate CSS file
library(xaringanthemer)
rcfss::xaringan_theme()

# source in the default knitr options
source(here::here("R", "slide-opts.R"))

# enable panelsets and default theme
xaringanExtra::use_panelset()
# ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# load basic packages
library(knitr)
library(here)
library(countdown)
library(patchwork)
library(flipbookr)
library(scales)
```

## Interactive visualizations with Shiny

We will build an app with

- Data pre-processing
- Dynamic UI elements
- Linked brushing

---

class: center, middle

[https://bensoltoff.shinyapps.io/manager-survey/](https://bensoltoff.shinyapps.io/manager-survey/)

```{r echo = FALSE}
knitr::include_app("https://bensoltoff.shinyapps.io/manager-survey/", height ="550px")
```

---

class: middle

.large[.hand[livecoding]]

.task[
Go to the `shiny` project and code along in `manager-survey/app.R`.

Make sure you are starting with the code [here](https://github.com/uc-dataviz/slides/blob/main/15-interactive-reporting/manager-survey/app.R).

]

(See next slide for code developed during live coding session)

---

class: middle

The code for the app can be found [here](https://github.com/uc-dataviz/slides/blob/main/16-interactive-visualization/manager-survey/app.R).
